---
title: "Day 4: a bad map"
author: "Arthur Bazolli"
date: 2023-11-04
---

So here's the deal: imma make a face out of Minas Gerais, manually painting cities. 

```{r}
#| label: setup

library(sf)
library(geobr)
library(tmap)
library(tidyverse)

tmap_options(check.and.fix = T)
tmap_mode("view")
```


# Download data

Once again, I really love ipeaGIT!

```{r}
#| label: download

shp_mg <- read_state(code_state = "31")
shp_muni <- read_municipality(code_muni = "MG")
shp_ri <- read_immediate_region(code_immediate = "MG")
```



# Visual inspection

The first step is to navigate the city. And here's my trick: I use tmap's interactive mode! When you hover over a city, it'll display its IBGE code, which I put on the lists below. 
```{r}
#| label: tmap_v1

tm_shape(shp_muni) +
  tm_fill(col = "grey") +
  tm_borders()
```


I'll skip all my painful iteractions and just show you the final maps. Final draft, using tmap:
```{r}
#| label: tmap_v2

tm_shape(shp_muni) +
tm_fill(col = "lightgrey") +
tm_borders() +
tm_shape(shp_muni %>% filter(code_muni %in% eye)) +
tm_fill(col = "black") +
tm_shape(shp_muni %>% filter(code_muni %in% eyebrow)) +
tm_fill(col = "#543808") +
tm_shape(shp_ri %>% filter(code_immediate %in% hair)) +
tm_fill(col = "#7d5819") +
tm_shape(shp_muni %>% filter(code_muni %in% teeth)) +
tm_fill(col = "#EAE0C8") +
tm_borders(col = "black") +
tm_shape(shp_muni %>% filter(code_muni %in% mouth)) +
tm_fill(col = "#d32f2f") +
tm_shape(shp_muni %>% filter(code_muni %in% booger)) +
tm_fill(col = "#b2d63f")
```



# Final map

Voilà, `ggplot`ed!
```{r}
#| label: final_map

ggplot() +
geom_sf(data = shp_mg, fill = "lightgrey", color = "black") +
geom_sf(data = shp_muni %>% filter(code_muni %in% eye), aes(fill = "eye"), color = "black") +
geom_sf(data = shp_muni %>% filter(code_muni %in% eyebrow),
aes(fill = "eyebrow"), color = "#543808") +
geom_sf(data = shp_ri %>% filter(code_immediate %in% hair),
aes(fill = "hair"), color = "#7d5819") +
geom_sf(data = shp_muni %>% filter(code_muni %in% mouth),
aes(fill = "mouth"), color = "#d32f2f") +
geom_sf(data = shp_muni %>% filter(code_muni %in% booger),
aes(fill = "booger"), color = "#b2d63f") +
geom_sf(data = shp_muni %>% filter(code_muni %in% teeth),
aes(fill = "teeth"), color = "black") +
scale_fill_manual(values = c("eye" = "black", "eyebrow" = "#543808", "hair" = "#7d5819",
"teeth" = "#EAE0C8", "mouth" = "#d32f2f", "booger" = "#b2d63f"),
guide = "legend") +
theme_void() +
labs(title = "minas gerais as a person", fill = "parts") +
theme(
text = element_text(family = "Comic Sans MS")
)
```

